A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-05-28 main.tex
Translation table: paper symbol to Abella relation
Some revisions

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

@@

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been \emph{manually} proven
\emph{sound}, \emph{complete} and \emph{decidable}. We present 
a mechanical formalization in the Abella theorem prover 
of DK's declarative type system with a novel algorithmic system. We have
a few reasons to use a new algorithm. Firstly, our new algorithm
employs \emph{worklist judgments}, which allow precise capture of the
scope of variables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing and some
incorrect proofs,
which complicate the task of writing a machanized proof. 
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. Since such type inference algorithms are quite
subtle and have a complex metatheory, mechanical
formalizations are an important advance in type-inference research.
\end{abstract}

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max
%% submission space)
\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

\usepackage{pdflscape}
\usepackage{afterpage}
% \usepackage{rotating}

@@

This paper presents the first full mechanical formalization of the metatheory for
higher-ranked polymorphic type inference. The system that we formalize
is the bidirectional type system by Dunfield and
Krishnaswami (DK). The DK type system has two variants (a declarative
and an algorithmic one) that have been \emph{manually} proven
\emph{sound}, \emph{complete} and \emph{decidable}. We present 
a mechanical formalization in the Abella theorem prover 
of DK's declarative type system with a novel algorithmic system. We have
a few reasons to use a new algorithm. Firstly, our new algorithm
employs \emph{worklist judgments}, which precisely capture the
scope of variables and simplify the formalization of scoping in a
theorem prover. Secondly, while DK's original formalization comes with very
well-written manual proofs, there are several details missing and some
incorrect proofs,
which complicate the task of writing a mechanized proof. 
Despite the use of a different algorithm we prove the
same results as DK, although with significantly different proofs and
proof techniques. Since such type inference algorithms are quite
subtle and have a complex metatheory, mechanical
formalizations are an important advance in type-inference research.
\end{abstract}

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.

We sincerely thank the anonymous reviewers for their insightful
comments. Ningning Xie found the issue with Lemma 29 that we reported 
on this article. This work has been sponsored by the Hong Kong Research
Grant Council projects number 17210617 and 17258816, and by the
Research Foundation - Flanders.
\end{acks}

### 2019-05-28 sections/Overview.tex
minor typos
Some revisions

@@

\paragraph{Declarative Subtyping}
Figure~\ref{fig:decl:sub} shows DK's declarative subtyping judgment $\Psi \vdash A \le B$,
which was adopted from \citet{odersky1996putting}. It compares the
degree of polymorphism between $A$ and $B$ in DK's implicit polymorphic type system. 
Essentially, if $A$ can always be instantiated to match any instantiation of $B$,
then A is ``at least as polymorphic as'' $B$. We also 
say that $A$ is ``more polymorphic than'' $B$ and write $A \le B$.

@@

\jimmy{Mention the implicit freshness conditions in the premises throughout the text}
Please note that when we introduce a new binder in the premise, we implicitly pick a fresh one.
This applies to rules such as $\mathtt{wf_d\forall}$, $\mathtt{wf_dabs}$, $\mathtt{{\le}\forall R}$,
throughout the whole text.
For algorithmic rules in Figure~\ref{fig:alg}, the right-hand side $\Gm'$
is actually the premise if re-written in bit-step style.

@@

Many rules are standard.
Rule $\mathtt{DeclVar}$ looks up term variables in the context.
Rules $\mathtt{Decl1I}$ and $\mathtt{Decl1I{\To}}$ respectively check and synthesize the unit type.
Rule $\mathtt{DeclAnno}$ synthesizes the annotated type $A$ of the annotated expression $(e:A)$
and checks that $e$ has type $A$.
Checking an expression $e$ against a polymorphic type $\all A$ in the context $\Psi$ succeeds
if $e$ checks against $A$ in the extended context $(\Psi, a)$.
The subsumption rule $\mathtt{DeclSub}$ depends on the subtyping relation,
and changes mode from checking to synthesis: if $e$ synthesizes type $A$ and $A\le B$
($A$ is more polymorphic than $B$), then $e$ checks against $B$.
If a checking problem does not match any other rules,
this rule can be applied to synthesize a type instead and then
check whether the synthesized type entails the checked type.
Lambda abstractions are the hardest construct of the bidirectional
type system to deal with. 
Checking $\lam e$ against function type $A\to B$ is easy:
we check the body $e$ against $B$ in the context extended with $x:A$.
However, synthesizing a lambda-function is a lot less easy, and 
this type system only synthesizes monotypes $\sigma\to\tau$.

@@

\paragraph{Overlapping Rules}
Some of the declarative rules overlap with each other.
Declarative subtyping rules $\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$
both match the conclusion $\Psi\vdash \all A \le \all B$.
In such case, choosing $\mathtt{{\le}\forall R}$ first is always better,
since we introduce the type variable $a$ to the context earlier,
which gives more flexibility on the choice of $\tau$.
Declarative typing rule $\mathtt{DeclSub}$ overlaps with
both $\mathtt{Decl\forall I}$ and $\mathtt{Decl{\to}I}$.
However, we argue that more specific rules are always the best choices,
i.e. $\mathtt{Decl\forall I}$ and $\mathtt{Decl{\to}I}$ should have
higher priority than $\mathtt{DeclSub}$. We will come back to this
topic in Section~\ref{sec:metatheory:non-overlapping}.
\begin{comment}
For example, $\Psi\vdash \lam x \Lto \all a\to a$ succeeds if derived from
Rule $\mathtt{Decl\forall I}$, but fails when applied to $\mathtt{DeclSub}$:
$$
\inferrule*[right={$\mathtt{Decl\forall I}$}]
	{\inferrule*[Right={$\mathtt{Decl\to I}$}]
		{\Psi,a,x:a \vdash x \Lto a}
		{\Psi,a \vdash \lam x \Lto a \to a}
	}
	{\Psi\vdash \lam x \Lto \all a \to a}
$$
$$
\inferrule*[right={$\mathtt{DeclSub}$}]
	{
		\inferrule*[right=$\mathtt{Decl\to I\To}$]
			{\Psi \vdash \blue\sigma\to \blue\tau \quad \Psi,x:\blue\sigma\vdash e \Lto \blue\tau}
			{\Psi \vdash \lam x \To \blue\sigma\to \blue\tau}\quad
		\inferrule*[Right=$\mathtt{{\le}\forall R}$]
			{\inferrule*[Right=$\mathtt{{\le}{\to}}$]
				{
					\inferrule*[Right=$?$]
						{\text{Impossible! }\blue\sigma \neq a}
						{\Psi,a \vdash a \le \blue\sigma}
					\quad \Psi,a \vdash \blue\tau \le a
				}
				{\Psi,a \vdash \blue\sigma\to \blue\tau \le a \to a}
			}
			{\Psi\vdash \blue\sigma\to \blue\tau\le \all a \to a}
	}
{\Psi\vdash \lam x \Lto \all a \to a}
$$

@@

More seriously, we have also found a lemma that simply does not hold, 
Lemma 29 (Parallel Admissibility).
See Appendix~\ref{appendix:false_lemmas} for a detailed explanation
and counterexample. This lemma is a cornerstone of the two metatheoretical results 
of the algorithm, soundness and completeness with respect to the declarative system.
In particular, both instantiation soundness (i.e. a part of subtyping
soundness) and typing completeness directly require the broken lemma.
Moreover, Lemma 54 (Typing Extension) also requires the broken lemma and is
itself used 13 times in the proof of typing soundness and completeness.
Unfortunately, we have not yet found a way to fix this problem.
\footnote{We found the issue with Lemma 29 in 2016 in an earlier attempt to mechanically
  formalize DK's algorithm. The authors acknowledged the problem after we contacted them through email.
  Although they briefly mentioned that it should be possible to use a weaker lemma instead they did
  not go into details.
}

@@

The complications of scoping and propagation are compelling reasons
to look for a simpler algorithm that is easier to
reason about in a mechanized setting.

%-------------------------------------------------------------------------------
\subsection{Judgment Lists}\label{sec:overview:list}

To avoid the problem of incrementally applying a substitution to remaining
tasks, we can find inspiration in the formulation of constraint solving
algorithms. For instance, the well-known unification
algorithm by \citet{unification} decomposes the problem of unifying two terms $s \stackrel{.}{=} t$ into a number
of related unification problems between pairs of terms $s_i \stackrel{.}{=} t_i$. These smaller
problems are not tackled independently, but kept together in a set $S$. 
The algorithm itself is typically formulated as a small-step-style state
transition system $S \rightarrowtail S'$ that rewrites the set of unification
problems until it is in solved form or until a contradiction has been found.
For instance, the variable elimination rule is written as:
\[
   x \stackrel{.}{=} t, S  ~~\rightarrowtail~~  x \stackrel{.}{=} t, [t/x]S   \qquad\qquad{if}~x \not\in t~\text{and}~{x \in S}
\]
Because the whole set is explicitly available, the variable $x$ can be
simultaneously substituted everywhere.

@@

In recent work, \citet{itp2018} have shown how to mechanize a variant of DK's
subtyping algorithm and shown it correct with respect to DK's declarative
subtyping judgment. This approach overcomes the some problems in DK's formulation
by using a \emph{worklist}-based judgment of the form $$\Gamma \vdash \Omega$$
where $\Omega$ is a worklist (or sequence) of subtyping problems of the
form $A \leq B$.  The judgment is defined by case analysis on the first
element of $\Omega$ and recursively processes the worklist until it is empty.
Using both a single common ordered context $\Gamma$ and a worklist $\Omega$ greatly
simplifies propagating the instantiation of type variables in one
subtyping problem to the remaining ones.

>>>>

\paragraph{Well-formedness} Well-formedness of types and terms is 
shown at the top of Figure~\ref{fig:decl:sub}. The rules are standard
and simply ensure that variables in types and terms are well-scoped.

@@

\paragraph{Declarative Subtyping}
The bottom of Figure~\ref{fig:decl:sub} shows DK's declarative subtyping judgment $\Psi \vdash A \le B$,
which was adopted from \citet{odersky1996putting}. It compares the
degree of polymorphism between $A$ and $B$ in DK's implicit polymorphic type system. 
Essentially, if $A$ can always be instantiated to match any instantiation of $B$,
then A is ``at least as polymorphic as'' $B$. We also 
say that $A$ is ``more polymorphic than'' $B$ and write $A \le B$.

@@

\jimmy{Mention the implicit freshness conditions in the premises throughout the text}
Please note that when we introduce a new binder in the premise, we implicitly pick a fresh one.
This applies to rules such as $\mathtt{wf_d\forall}$, $\mathtt{wf_dabs}$, $\mathtt{{\le}\forall R}$,
throughout the whole text.
For algorithmic rules in Figure~\ref{fig:alg}, the right-hand side $\Gm'$
is actually the premise if re-written in a big-step style.
\bruno{Comment about big-step seems out of place. Should come later,
  when we actually talk about the algorithm.}

@@

Many rules are standard.
Rule $\mathtt{DeclVar}$ looks up term variables in the context.
Rules $\mathtt{Decl1I}$ and $\mathtt{Decl1I{\To}}$ respectively check and synthesize the unit type.
Rule $\mathtt{DeclAnno}$ synthesizes the annotated type $A$ of the annotated expression $(e:A)$
and checks that $e$ has type $A$.
Checking an expression $e$ against a polymorphic type $\all A$ in the context $\Psi$ succeeds
if $e$ checks against $A$ in the extended context $(\Psi, a)$.
The subsumption rule $\mathtt{DeclSub}$ depends on the subtyping relation,
and changes mode from checking to synthesis: if $e$ synthesizes type $A$ and $A\le B$
($A$ is more polymorphic than $B$), then $e$ checks against $B$.
If a checking problem does not match any other rules,
this rule can be applied to synthesize a type instead and then
check whether the synthesized type entails the checked type.
Lambda abstractions are the hardest construct of the bidirectional
type system to deal with. 
Checking $\lam e$ against function type $A\to B$ is easy:
we check the body $e$ against $B$ in the context extended with $x:A$.
However, synthesizing a lambda-function is a lot harder, and 
this type system only synthesizes monotypes $\sigma\to\tau$.

@@

\paragraph{Overlapping Rules}
Some of the declarative rules overlap with each other.
Declarative subtyping rules $\mathtt{{\le}\forall L}$ and $\mathtt{{\le}\forall R}$
both match the conclusion $\Psi\vdash \all A \le \all B$.
In such case, choosing $\mathtt{{\le}\forall R}$ first is always better,
since we introduce the type variable $a$ to the context earlier,
which gives more flexibility on the choice of $\tau$.
The declarative typing rule $\mathtt{DeclSub}$ overlaps with
both $\mathtt{Decl\forall I}$ and $\mathtt{Decl{\to}I}$.
However, we argue that more specific rules are always the best choices,
i.e. $\mathtt{Decl\forall I}$ and $\mathtt{Decl{\to}I}$ should have
higher priority than $\mathtt{DeclSub}$. We will come back to this
topic in Section~\ref{sec:metatheory:non-overlapping}.
\begin{comment}
For example, $\Psi\vdash \lam x \Lto \all a\to a$ succeeds if derived from
Rule $\mathtt{Decl\forall I}$, but fails when applied to $\mathtt{DeclSub}$:
$$
\inferrule*[right={$\mathtt{Decl\forall I}$}]
	{\inferrule*[Right={$\mathtt{Decl\to I}$}]
		{\Psi,a,x:a \vdash x \Lto a}
		{\Psi,a \vdash \lam x \Lto a \to a}
	}
	{\Psi\vdash \lam x \Lto \all a \to a}
$$
$$
\inferrule*[right={$\mathtt{DeclSub}$}]
	{
		\inferrule*[right=$\mathtt{Decl\to I\To}$]
			{\Psi \vdash \blue\sigma\to \blue\tau \quad \Psi,x:\blue\sigma\vdash e \Lto \blue\tau}
			{\Psi \vdash \lam x \To \blue\sigma\to \blue\tau}\quad
		\inferrule*[Right=$\mathtt{{\le}\forall R}$]
			{\inferrule*[Right=$\mathtt{{\le}{\to}}$]
				{
					\inferrule*[Right=$?$]
						{\text{Impossible! }\blue\sigma \neq a}
						{\Psi,a \vdash a \le \blue\sigma}
					\quad \Psi,a \vdash \blue\tau \le a
				}
				{\Psi,a \vdash \blue\sigma\to \blue\tau \le a \to a}
			}
			{\Psi\vdash \blue\sigma\to \blue\tau\le \all a \to a}
	}
{\Psi\vdash \lam x \Lto \all a \to a}
$$

@@

More seriously, we have also found a lemma that simply does not hold:
Lemma 29 (Parallel Admissibility)\footnote{Ningning Xie found the issue with Lemma 29 in 2016 while
  collaborating with the second author on an earlier attempt to mechanically
  formalize DK's algorithm. The authors acknowledged the problem after we contacted them through email.
  Although they briefly mentioned that it should be possible to use a weaker lemma instead they did
  not go into details.}.
See Appendix~\ref{appendix:false_lemmas} for a detailed explanation
and counterexample. This lemma is a cornerstone of the two metatheoretical results 
of the algorithm, soundness and completeness with respect to the declarative system.
In particular, both instantiation soundness (i.e. a part of subtyping
soundness) and typing completeness directly require the broken lemma.
Moreover, Lemma 54 (Typing Extension) also requires the broken lemma and is
itself used 13 times in the proof of typing soundness and completeness.
Unfortunately, we have not yet found a way to fix this problem.

@@

The complications of scoping and propagation are compelling reasons
to look for another algorithm that is easier to
reason about in a mechanized setting.

%-------------------------------------------------------------------------------
\subsection{Judgment Lists}\label{sec:overview:list}
\bruno{Somewhere we should mention the ``Essence of ML type
  inference'' as pointed out by our reviews/comments?}
To avoid the problem of incrementally applying a substitution to remaining
tasks, we can find inspiration in the formulation of constraint solving
algorithms. For instance, the well-known unification
algorithm by \citet{unification} decomposes the problem of unifying two terms $s \stackrel{.}{=} t$ into a number
of related unification problems between pairs of terms $s_i \stackrel{.}{=} t_i$. These smaller
problems are not tackled independently, but kept together in a set $S$. 
The algorithm itself is typically formulated as a small-step-style state
transition system $S \rightarrowtail S'$ that rewrites the set of unification
problems until it is in solved form or until a contradiction has been found.
For instance, the variable elimination rule is written as:
\[
   x \stackrel{.}{=} t, S  ~~\rightarrowtail~~  x \stackrel{.}{=} t, [t/x]S   \qquad\qquad{if}~x \not\in t~\text{and}~{x \in S}
\]
Because the whole set is explicitly available, the variable $x$ can be
simultaneously substituted everywhere.

@@

In recent work, \citet{itp2018} have shown how to mechanize a variant of DK's
subtyping algorithm and shown it correct with respect to DK's declarative
subtyping judgment. This approach overcomes some problems in DK's formulation
by using a \emph{worklist}-based judgment of the form $$\Gamma \vdash \Omega$$
where $\Omega$ is a worklist (or sequence) of subtyping problems of the
form $A \leq B$.  The judgment is defined by case analysis on the first
element of $\Omega$ and recursively processes the worklist until it is empty.
Using both a single common ordered context $\Gamma$ and a worklist $\Omega$ greatly
simplifies propagating the instantiation of type variables in one
subtyping problem to the remaining ones.

### 2019-05-28 sections/algorithm.tex
Some revisions

Figure~\ref{fig:alg:syntax} shows the syntax and well-formedness judgments used by the algorithm.

@@

\paragraph{Worklist Variable Scoping}
Rules 8 and 9 involve variable declarations and demonstrate how our
worklist treats variable scopes. Rule 8 introduces an existential
variable $\al$ that is only visible to the judgment $[\al/a]A \le B$.
Reduction continues until all the subtyping judgments in front of $\al$
are satisfied.  Finally we can safely remove $\al$ since no occurrence
of $\al$ could have leaked into the left part of the worklist.  Moreover,
the algorithm garbage-collects the $\al$
variable at the perfect time: it leaves the environment immediately
after being unreferenced completely for sure.

>>>>

Figure~\ref{fig:alg:syntax} shows the syntax and well-formedness
judgments used by the algorithm. Similarly to the declarative system 
the well-formedness rules are unsurprising and merely ensure 
well-scoping of variables.

@@

\paragraph{Worklist Variable Scoping}
Rules 8 and 9 involve variable declarations and demonstrate how our
worklist treats variable scopes. Rule 8 introduces an existential
variable $\al$ that is only visible to the judgment $[\al/a]A \le B$.
Reduction continues until all the subtyping judgments in front of $\al$
are satisfied.  Finally we can safely remove $\al$ since no occurrence
of $\al$ could have leaked into the left part of the worklist.  Moreover,
the algorithm garbage-collects the $\al$
variable at the right time: it leaves the environment immediately
after being unreferenced completely for sure.

### 2019-05-28 sections/introduction.tex
Some revisions

This paper presents the first full mechanical formalization of the
metatheory for higher-ranked polymorphic type inference.
The system
that we formalize is the bidirectional type system by \citet{dunfield2013complete}.
We chose DK's type system because it is
quite elegant, well-documented and it comes with detailed manually
written proofs. Furthermore the system is adopted in practice by a few
real implementations of functional languages, including PureScript and
Unison. The DK type system has two variants: a declarative
and an algorithmic one. The two variants have been
\emph{manually} proved to be \emph{sound}, \emph{complete} and
\emph{decidable}.
We present a mechanical formalization in the Abella theorem prover~\cite{AbellaDesc} for
DK's declarative type system using a different algorithm. While our
initial goal was to formalize both DK's declarative and algorithmic
versions, we faced technical challenges with the latter, prompting us to find
an alternative formulation.

@@

\item {\bf A full mechanical formalization of type inference with
  higher-ranked types:} Our work presents the first full mechanical formalization
  for type inference of higher ranked types. The formalization is done in the
  Abella theorem prover~\cite{AbellaDesc} and it is available
  online.\footnote{{\bf Note to reviewers:} Due to the anonymous submission process
  it is sent as part of the supplementary materials.}

>>>>

This paper presents the first fully mechanized formalization of the
metatheory for higher-ranked polymorphic type inference.
The system
that we formalize is the bidirectional type system by \citet{dunfield2013complete}.
We chose DK's type system because it is
quite elegant, well-documented and it comes with detailed manually
written proofs. Furthermore the system is adopted in practice by a few
real implementations of functional languages, including PureScript and
Unison. The DK type system has two variants: a declarative
and an algorithmic one. The two variants have been
\emph{manually} proved to be \emph{sound}, \emph{complete} and
\emph{decidable}.
We present a mechanical formalization in the Abella theorem prover~\cite{AbellaDesc} for
DK's declarative type system using a different algorithm. While our
initial goal was to formalize both DK's declarative and algorithmic
versions, we faced technical challenges with the latter, prompting us to find
an alternative formulation.

@@

\item {\bf A fully mechanized formalization of type inference with
  higher-ranked types:} Our work presents the first fully mechanized formalization
  for type inference of higher ranked types. The formalization is done in the
  Abella theorem prover~\cite{AbellaDesc} and it is available
  online at \url{http://TODO}.\bruno{fill url please!}
